---
title: "Split usefulness evaluation (2018-19)"
author: "CourtSightR"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(knitr)
# If rendered from reports/, go to project root so paths to scripts/ and reports/tables/ work
if (basename(getwd()) == "reports") setwd("..")
```

# Question

Which split types in CourtSightR most consistently produce **meaningful, non-trivial, and reasonably stable variation** for player-level analysis after filtering out tiny or noisy groups?

This is a **practical analytics question** (which splits are useful for exploratory analysis), not a claim about basketball causality or universal truth.

# Dataset

- **Source:** `data/games_2018_19_courtsightr.csv` (cleaned 2018-19 NBA player-game data).
- **Scope:** The slice uses a **date window** (2018-10-01 to 2019-06-30). The source does **not** provide a playoff flag, so the slice includes **both regular season and playoffs**.
- **Entity filter:** Top **30** players by total minutes, with **≥ 45 games** and **≥ 20 average minutes** per game. This keeps splits interpretable and avoids noise from low-usage players.

# Methods

- **Binary splits** (`win_loss`, `home_away`): For each player and metric, we require **≥ 10 games** in each of the two groups. We compute **Cohen’s d** (standardized mean difference) between groups, then take the **median of |Cohen’s d|** across qualifying players as the usefulness score. **Uncertainty:** 95% bootstrap interval for that median (1000 reps).
- **Multi-group split** (`opponent`): For each player we require **≥ 5 games** per opponent and **≥ 4 opponent groups** and **≥ 30 total rows**. We compute **eta-squared** (variance explained by opponent) per player, then **median eta-squared** across players. **Uncertainty:** 95% bootstrap interval for that median.
- **Metrics:** points, rebounds, assists, plus_minus, and **fg_pct** (derived as fg_made/fg_attempts per game when fg_attempts > 0).
- **Useful:** For binary splits we interpret |Cohen’s d| as trivial (&lt; 0.20), small (0.20–0.50), moderate (0.50–0.80), or large (≥ 0.80). For opponent we use eta² cutoffs: trivial (&lt; 0.01), small (0.01–0.06), moderate (0.06–0.14), large (≥ 0.14).

# Results

Run the evaluation script once to generate the summary table:

```bash
Rscript scripts/evaluate_split_usefulness.R
```

Then the table below is read from `reports/tables/split_usefulness_summary.csv`.

```{r load-table}
csv_path <- file.path("reports", "tables", "split_usefulness_summary.csv")
if (!file.exists(csv_path)) {
  stop("Run first from project root: Rscript scripts/evaluate_split_usefulness.R")
}
tbl <- read_csv(csv_path, show_col_types = FALSE)
```

## Comparison table

```{r comparison-table}
kable(tbl, digits = 4)
```

# Failure cases / misleading split examples

- **Binary:** Players with &lt; 10 games in either group (e.g. win vs loss) are excluded, so the score reflects only players with enough data in both groups.
- **Opponent:** Imbalanced schedules (e.g. few games vs some teams) can inflate or deflate eta²; we mitigate by requiring at least 5 games per opponent and 4+ opponent groups.
- Splits that show “insufficient data” or “trivial” in the table indicate that for this season and player pool, that split did not yield a stable or meaningful signal.

# Limitations

- **2018-19 slice includes playoffs**; we do not separate regular season from playoffs.
- **Source data** is scraped (e.g. Basketball Reference–style), not official; no guarantee of completeness or accuracy.
- **No causal claims**; we only assess whether split types show consistent variation, not whether the split “causes” performance differences.
- **Single season, single player pool**; results are not a universal validation of the tool.

# Conclusion

The comparison table summarizes which split × metric combinations produced non-trivial, stable variation in this evaluation. Use it to prioritize which splits are most informative for exploratory player-level analysis on the 2018-19 cleaned dataset. The tool remains a **local-first analytics helper**; this evaluation is a small, defensible check of split usefulness, not a full validation.
