---
title: "Split usefulness evaluation (2018-19)"
author: "CourtSightR"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(knitr)
# If rendered from reports/, go to project root so paths to scripts/ and reports/tables/ work
if (basename(getwd()) == "reports") setwd("..")
```

# Scope and framing

This is an **exploratory** evaluation of which CourtSightR split types produce meaningful, stable variation for player-level analysis. It is not a causal study, a prediction project, or a formal validation of the tool.

- **Data:** The 2018-19 slice includes **both regular season and playoffs** (date window only; the source has no playoff flag).
- **Source:** Scraped (e.g. Basketball Reference–style), not official; no guarantee of completeness or accuracy.
- **Findings:** They describe **practical split usefulness** for exploratory analysis, not causation or universal basketball truth.
- **Opponent split:** Only a **limited subset of players** (8 of 30 in this run) met the opponent-group thresholds; opponent results should be interpreted cautiously.

# Question

Which split types in CourtSightR most consistently produce **meaningful, non-trivial, and reasonably stable variation** for player-level analysis after filtering out tiny or noisy groups?

# Dataset

- **Source:** `data/games_2018_19_courtsightr.csv` (cleaned 2018-19 NBA player-game data).
- **Scope:** Date window 2018-10-01 to 2019-06-30. The source does **not** provide a playoff flag, so the slice includes **both regular season and playoffs**.
- **Player pool:** Top **30** players by total minutes, with **≥ 45 games** and **≥ 20 average minutes** per game.

# Methods

- **Binary splits** (`win_loss`, `home_away`): For each player and metric, **≥ 10 games** in each of the two groups. Use **Cohen’s d** between groups, then **median of |Cohen’s d|** across qualifying players as the usefulness score. **Uncertainty:** 95% bootstrap interval for that median (1000 reps).
- **Multi-group split** (`opponent`): For each player, **≥ 5 games** per opponent, **≥ 4 opponent groups**, **≥ 30 total rows**. Use **eta-squared** (variance explained by opponent) per player, then **median eta-squared** across qualifying players. **Uncertainty:** 95% bootstrap interval for that median. In this run, **only 8 players** met these thresholds; opponent results are exploratory for that smaller subset.
- **Metrics:** points, rebounds, assists, plus_minus, fg_pct (derived as fg_made/fg_attempts per game when fg_attempts > 0).
- **Interpretation:** Binary: |Cohen’s d| trivial (&lt; 0.20), small (0.20–0.50), moderate (0.50–0.80), large (≥ 0.80). Opponent: eta² trivial (&lt; 0.01), small (0.01–0.06), moderate (0.06–0.14), large (≥ 0.14).

# Results

The table below is read from `reports/tables/split_usefulness_summary.csv` (pre-computed in the repo).

```{r load-table}
csv_path <- file.path("reports", "tables", "split_usefulness_summary.csv")
if (!file.exists(csv_path)) {
  stop("Summary table not found. Run from project root: Rscript scripts/evaluate_split_usefulness.R")
}
tbl <- read_csv(csv_path, show_col_types = FALSE)
```

## Comparison table

```{r comparison-table}
kable(tbl, digits = 4)
```

## Top-line findings (from this run)

- **win_loss** is the **strongest overall split** in this evaluation. Median |Cohen’s d|: **plus_minus 1.76** (large; 95% bootstrap CI 1.66–1.85), **assists 0.33** and **fg_pct 0.42** (small), **points 0.21** (small), **rebounds 0.18** (trivial). All 30 players qualified. **Note:** Large plus_minus differences under win_loss are expectedly tied to outcome (wins vs losses) and should not be treated as a surprising insight.
- **home_away** is **weaker and usually small**. Median |Cohen’s d|: **plus_minus 0.42** (small), all other metrics trivial (points 0.15, rebounds 0.17, assists 0.13, fg_pct 0.16). All 30 players qualified.
- **opponent** showed **moderate signal for a smaller qualifying subset only**. Only **8 players** met the opponent thresholds. Median eta²: **plus_minus 0.16** (large explained variation), **points 0.07**, **rebounds 0.10**, **assists 0.08**, **fg_pct 0.09** (moderate). These results are exploratory and should be interpreted cautiously given the small number of qualifying players and schedule/opponent imbalance.

# Failure cases / caveats

- **Binary:** Players with &lt; 10 games in either group are excluded.
- **Opponent:** Imbalanced schedules can inflate or deflate eta²; only 8 of 30 players had enough opponent diversity for inclusion.
- “Trivial” or “insufficient data” in the table means that split did not yield a stable or meaningful signal in this run.

# Limitations

- **2018-19 slice includes playoffs**; regular season is not separated.
- **Source data** is scraped, not official.
- **No causal claims**; we only assess whether split types show consistent variation.
- **Single season, single player pool**; results are not a universal validation of the tool.
- **Opponent** results are based on 8 qualifying players only.

# Conclusion

**win_loss** is the strongest split in this exploratory evaluation; **home_away** is weaker and usually small; **opponent** shows moderate signal for a limited subset. Use the table to prioritize which splits are most informative for exploratory player-level analysis on the 2018-19 cleaned dataset. The tool remains a local-first analytics helper; this evaluation is a small, defensible check of split usefulness, not a full validation.
